---
title: "R Notebook"
output: html_notebook
---


```{r}
# phylogenetic comparative analysis
library(ggtree)
library(ape)
library(phytools)
library(dplyr)
library(readxl)
library(stringr)
library(ggthemr)
library(reshape2)
library(scales)
library(forcats)
library(readr)
# for comparative analysis
library(caper)
library(yhat)
library(dplyr)
library(GGally)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

Load the data and do some processing.

```{r, results = 'hide'}
# phylogeny
tree_final <- read.tree("data/raw/phylogeny/higdon_mod2_28.tre")

# all other seal stats and demographic variables
all_stats <- read_csv("data/processed/all_stats_tree.csv") %>% 
    mutate(SSD = male_weight/female_weight) %>% 
    mutate(abc_out = ifelse(bot > 0.5, "bot", "neut")) %>% 
    mutate(BreedingType = factor(BreedingType, levels = c("ice", "land", "both"))) %>% 
    mutate(logAbundance = log(Abundance),
        logharem_size = log(harem_size),
        logmale_weight = log(male_weight)) %>% 
    # order factors according to tree
    mutate(tip_label = fct_inorder(factor(tip_label)),
        species = fct_inorder(factor(species)),
        latin = fct_inorder(factor(latin)),
        common = fct_inorder(factor(common)),
        short = fct_inorder(factor(short)))
# count grey and harbour seal to land breeding
all_stats[all_stats$BreedingType == "both", "BreedingType"] <- "land" 
all_stats <- all_stats %>% mutate(BreedingType = as.factor(as.character(BreedingType))) %>% data.frame()

# produce short names for plotting
short <- c("W", "NFS", "SSL", "CSL", "GSL", "SASL", "AFS", "NZSL", "AntFS", "NZFS", "SAFS", "GFS", 
    "BS", "HoS", "GS", "HS", "ARS", "SRS", "BRS", "LRS", "MMS", "HMS", "NES", "SES", "CS", "RS", "LS", "WS")
all_stats$short <- short
```

Here is the phlogeny.

```{r, fig.height=7, fig.width=7}
plot(tree_final)
```


Select variables for modeling.

```{r}
stats_mod <- all_stats %>% 
                dplyr::select(TPM70_ratio, TPM90_ratio, num_alleles_mean, harem_size, SSD, BreedingType, 
                              male_weight, breeding_season_length, lactation_length, life_span_years, 
                              Abundance, Generation_time, Longevity, tip_label, mratio_mean,
                              obs_het_mean, mean_allele_range, IUCN_rating, prop_low_afs_mean,
                              nloc, nind, bot) %>% 
                              data.frame()
```


Phylogenetic mixed modeling: Inverse phylo matrix and define priors.

```{r}
library(MCMCglmm)
inv_phylo <- inverseA(tree_final, nodes="TIPS")$Ainv #,scale=TRUE
prior<-list(G=list(G1=list(V=1,nu=0.002)),R=list(V=1,nu=0.002))
```


Hypothesis (1) How is heterozygosity-excess as a bottleneck signature correlated to typical diversity indices?

First, standardize input variables to make the estimates comparable.
```{r}
stats_mod_div <- stats_mod %>% 
                    mutate(num_alleles_mean = scale(num_alleles_mean),
                           obs_het_mean = scale(obs_het_mean),
                           prop_low_afs_mean = scale(prop_low_afs_mean))
```

```{r}
mod_gen <- MCMCglmm(TPM70_ratio ~ num_alleles_mean +  obs_het_mean + prop_low_afs_mean, #
    random=~tip_label, nodes = "TIPS", #   rcov =~us(trait):units
    family=c("gaussian"),ginverse=list(tip_label=inv_phylo),prior=prior,
    data=stats_mod_div,nitt=110000,burnin=10000,thin=50, verbose = FALSE)
```

```{r}
summary(mod_gen)
```


```{r, echo = TRUE}
plot(mod_gen$Sol)
```

```{r}
plot(mod_gen$VCV)
```


```{r}
autocorr(mod_gen$Sol)
```


```{r}
autocorr(mod_gen$VCV)
```







